---
title:
output:
  html_document:
    includes:
      in_header: header.html
---

<style>
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=1000)
```

# Chapter 13: R Tutorial on Relational Event Models

Built with R version
`r getRversion()`

This is the third tutorial for Chapter 13 on statistical network models. The first tutorial covered the case of cross-sectional network data. The second tutorial covered statistical models for discrete, longitudinal networks. Here, we will walk through relational event models, appropriate for continuous-time network data. Relational event models are based on micro-interaction data. The model assumes that there is time-stamped (or at least ordered) information on the interactions between a set of actors. This shifts the focus from discrete relationships (friend, advice, etc.) to the specific interactions between actors in a setting. The goal of the model is to predict what the next event is likely to be, based on the interactional tendencies, or rules, of behavior in the setting. Compare this to STERGM, where the goal is to predict the adding/dropping of ties from one period to the next, based on discretely defined networks. 

Our data for this tutorial are based on streaming interaction data collected by Daniel McFarland on students in classrooms. Time-stamped interactions in each classroom were recorded, with information on the 'sender' and 'receiver' of the interaction, as well as the nature of the interaction. Interactions could be social or task-based, for example. Data were collected across a large number of classrooms and days. See also Chapter 3c (data processing for dynamic network data) and Chapter 5 (on visualization). Here we consider one classroom on two different days; both days are in the second semester of the year. We pick two days as a means of comparison. The first day was relatively uneventful and class was orderly. The second day was different, as there was a much higher rate of sanctioning behavior (i.e., disagreements between students and teacher on what was going on in the classroom). By examining two days, we see if the interactional signatures of order and disorder are different. 

### 1 Getting the Data Ready 
Let’s begin by loading the main packages and getting the data ready to run the models. 

```{r message=F}
library(relevent)
library(sna)
```

relevent contains the functions to run relational event models. We can cite relevent as: Carter T. Butts (2008). A Relational Event Framework for Social Action. Sociological Methodology, 38(1), 155--200.

Now, let's read in the interactional data for the first date. This is a data set reporting on the social interactions, i.e. talking, between individuals in the classroom. This will serve as the outcome of interest, as we will predict what interactional features make certain events more likely to occur than others. 

```{r}
interactions_date1=read.table("https://sites.google.com/site/jeffreysmithdatafiles/class_interactions_date1.txt", 
                              header=T)
```

Here we take a look at the first six rows of the data, for the main variables of interest: send_col (id of sender), receive_col (id of receiver), and time_estimate_col (time that interaction occurred).

```{r}
head(interactions_date1[,c("send_col", "receive_col", "time_estimate_col")])
```

Each row corresponds to an interaction between sender and receiver. For example, we can see that the third social interaction in this class period involved node 2 talking to node 5. 

We need to manipulate the data a bit to get it in a form that the relational event model can use. For example, the events must be sorted in sequential order before we run any models. In this case, this is already done, but if it were not we would have to sort our data appropriately. Additionally, each interaction must also happen at a unique time period. The models are based on sequences of interactions, so a distinct order of events must be possible to establish. This means that relational event models are not so appropriate in cases where there are a large number of simultaneous events. For the sake of simplicity, we will remove all interactions directed from the teacher to all students or from all students to the teacher.

We can use the to_all_col and from_all_col to exclude these cases. to_all_col is equal to 1 if node i broadcasts to all other nodes simultaneously, while from_all_col is equal to 1 if node j receives from all nodes simultaneously. We will only keep those interactions where those variables are equal to 0 (i.e., i is not broadcasting to everyone in the class).

```{r}
interactions_date1=interactions_date1[interactions_date1$to_all_col==0 & interactions_date1$from_all_col==0,]
```

Now, in order to run the model, we need to create an edgelist (as a matrix object), where the first column is the time of the event, the second column is the sender and the third column is the receiver. Again, the events must be ordered sequentially. 

```{r}
edgelist_date1=as.matrix(interactions_date1[,c("time_estimate_col", "send_col", "receive_col")])
```

As a final data manipulation, we need to add a row to the end of the edgelist, showing the stop time where no more interactions are possible. Let's look at the end of the data frame: 

```{r}
tail(edgelist_date1)
```

We can see that the last social interaction occurred at minute 43 in the class period. We will set the end of the interactional period at 43.10 (i.e., 6 seconds after the final interaction).  To do this we add a row to the end of the edgelist, with the end time and then two NA values (for sender and receiver). 

```{r}
edgelist_date1=rbind(edgelist_date1, c(43.10, NA, NA))
```

Now, we will read in some attribute data, as we want to use information on gender, grade, etc. as predictors in the model. 

```{r}
attributes=read.table("https://sites.google.com/site/jeffreysmithdatafiles/class_attributes.txt", 
                      header=T)
head(attributes)
```

There are four variables: id, gnd (gender: 1=male; 2=female); grd (grade: 10=10; 11=11; 16=teacher); rce (race: 3=Black; 4=White).

The relevent package requires that a researcher construct the node-level predictors as distinct columns (as opposed to using a factor command within the formula). So, we will recode our variables to create the desired dummy variables of interest. In this case, we will create a predictor for whether the node is a teacher or not and for gender. We also need to create a term for the intercept. We will utilize the recode function in the car package. 

```{r message=F}
library(car)
```

We first will create a variable for the intercept, which is a simple 1 for all nodes in the class. 

```{r}
attributes$intercept=1
```

Now, we create a variable called male that is a 0 if gnd is equal to 2 (female) and 1 if gnd is equal to 1 (male). 

```{r}
attributes$male=recode(attributes$gnd, as.factor=F, "c(1)=1; c(2)=0")
```

And here we do the same thing for grd, creating a binary variable called teacher. teacher is equal to 1 if they are a teacher (grd=16) and 0 otherwise. 

```{r}
attributes$teacher=recode(attributes$grd, as.factor=F, "c(16)=1; NA=NA; else=0")
```

Finally, it will also be useful to have the size of the class handy. We can calculate that as the number of rows in the attribute data frame. 

```{r}
class_size=nrow(attributes)
```

#### 1.2. Running Initial Models
Relational event modeling is based on the logic of hazard models (or event history models), where the model predicts the risk of an event occurring (i.e., the hazard) as a function of different kind of interactional terms. There are a number of terms that we can include, including terms for baseline node effects (e.g., girls interact more than boys). We can also include terms that capture more micro-dynamics. These are labeled p-shifts, or participation shifts, and are only based on the previous event in the sequence. For example, if A talks to B, then we might expect the very next event to be B talking to A. The model allows us to include these different kinds of terms as a means of seeing what rules govern the interactions in the case of interest. 

We are now in a position to run an initial relational event model. The function is rem.dyad. The main inputs are:

- edgelist=input edgelist in the form of time of event, sender, receiver
- n=size of network
- effects=vector with names of effects to be included in the model
- covar=list of covariates that must correspond to the terms specified in effects
- ordinal=T/F; T if data are ordinal (ordered but without specific time stamps); F if data include time specific information for each event

Our first model will be very simple and just includes an intercept, capturing the baseline rate for events to occur. We will set the effects to CovSnd. CovSnd is a basic sender effect, in this case initiating social interactions with others. We use the covar option to include specific terms for CovSnd. In this case we include the intercept (so all nodes are assumed to initiate interactions at the same rate). We set ordinal to FALSE as the data has time stamped information. Let's also set a seed to make it easier to replicate. 

```{r message=F, results='hide'}
set.seed(1000) 
mod1=rem.dyad(edgelist=edgelist_date1, n=class_size, effects=c("CovSnd"), 
                covar=list(CovSnd=attributes$intercept), ordinal=FALSE, hessian=TRUE)
```

And now we look at the results:

```{r}
summary(mod1)
```

The coefficient for the intercept (CovSnd.1) isn't all that interesting in itself, but it is important to understand what the coefficients mean (and how we can manipulate them) before moving to more complicated models. The first thing to note is that if we exponentiate this coefficient, we get the hazard of any event (interaction between i-j) occurring. Higher hazards mean the risk for an event occurring is higher. Second, if we multiple the hazard rate by the number of possible node pairs who could interact in a given moment, n*(n-1), we should get the expected number of interactions occurring per minute in the classroom. Finally, if we take the inverse of that (1/number of interactions per minute), we get the expected amount of time between events, or the wait time between events.  

```{r}
1/(18*17*exp(-4.057353))
```

The expected time between any event occurring is .189 standardized minutes (or .189*60=11.34 seconds). And let's check this against the real data. We will take the total number of minutes for that class and divide that by the total number of interactions that occurred. We will define the total number of minutes as 43.1, the end time set above. 

```{r}
total_classtime=43.1
```

We now define the total number of interactions. We will take the number of rows in the edgelist and subtract 1, as the last row is the stop time (not an interaction). 

```{r}
num_interations=nrow(edgelist_date1)-1 
time_between_events=total_classtime/num_interations
time_between_events
```

We can see the estimate from the model approximates the raw data quite well. We can do more substantively interesting things by incorporating the attributes of the nodes into the model. Let's first add a term for gender (coded as male=1 and female=0). We will add a sender effect, capturing whether males initiate fewer or greater interactions than females, as well as a receiver effect, capturing whether males receive fewer/greater interactions than females. 

Here we create two matrices, one for sending and one for receiving. Each matrix will include the covariates we want to include for the sending or receiving effects. We start with the sender covariate matrix, where we will include variables for the intercept (it still must be included) and male. 

```{r}
CovSnd1=cbind(attributes[,c("intercept", "male")])
```

And now we do the same thing for the receiver covariate matrix, including a variable for male (note that no intercept term is included here). Note that even though we already had male as a sender effect we need to include it separately as part of the receiver covariate matrix if we want to specify it as a receiver effect.

```{r}
CovRec1=cbind(attributes[,c("male")])
```

And now we are ready to estimate the model. The only difference from before is that we include "CovRec" in the effects and CovRec1 as part of the covariate (covar) list. 

```{r results='hide'}
mod2a=rem.dyad(edgelist_date1, n=class_size, effects=c("CovSnd", "CovRec"), 
                covar=list(CovSnd=CovSnd1, CovRec=CovRec1), ordinal=FALSE, hessian=TRUE)
```


```{r}
summary(mod2a)
```

We can see that the names of the variables are difficult to interpret. So, let's create a vector of more useful names, and put that on the outputted object. 

```{r}
coef.names=c("intercept", "Sender.male", "Receiver.male")
```

And now we put those names on the coef part of the rem object and redo the summary of the model.

```{r}
names(mod2a$coef)=coef.names
summary(mod2a)
```

The summarized output now has more easily interpretable names for the variables. Let's go ahead and interpret the coefficient on sender.male. We can start by interpreting the sender.male coefficient in terms of hazards. By exponentiating the coefficient, we get the relative hazard for males to initiate the next interaction compared to females. exp(-.601)=.548. This means that an event with males initiating has a hazard that is .584 times lower than an event with females initiating. 

Hazards themselves are bit hard to interpret. As an alternative, we can calculate mean wait times, or the expected time between events. Let's first calculate the expected time between two male-male events (where there is a male sender and a male receiver). Note that this calculation must incorporate all of the coefficients (intercept, Sender.male and Receiver.male). The calculation is the same as we did above with the intercept only model, where the expected wait is equal to: 1/(dyads_at_risk*hazard), where dyads_at_risk is the number of different ways that the event could occur, in this case the number of possible events that could involve two males. 

Let's do a quick table to see how many males are in the classroom.

```{r}
table(attributes$male)
```

We can see that there are 12 females and 6 males. This means that there are 6*5 different ways that we could have a boy as a sender and a boy as a receiver. We will use that in the calculation below: 

```{r}
dyads_at_risk=6*5
```

Now we calculate the hazard. We take the coefficients, multiply them by the vector of input values (here setting Sender.male and Receiver.male to 1), sum it up and then exponentiate it. 

```{r}
inputs=c(intercept=1, Sender.male=1, Receiver.male=1)
hazard_male_male=exp(sum(mod2a$coef*inputs))
hazard_male_male
```

And now to calculate wait time:

```{r}
1/(dyads_at_risk*hazard_male_male)
```

This means that we would expect to wait 4.14 minutes between events that involve two boys. Now, let's do the same thing for girl-girl interactions. Here we set Sender.male to 0 and Receiver.male to 0. 

```{r}
inputs=c(intercept=1, Sender.male=0, Receiver.male=0)
```

We define the dyads at risk to be 12*11 as there are 12 females in the class. 

```{r}
dyads_at_risk=12*11

```

Here we calculate the hazard. 

```{r}
hazard_female_female=exp(sum(mod2a$coef*inputs))
hazard_female_female
```

We can see that the hazard for female-female events is higher than male-male events. And now for the wait time.

```{r}
1/(dyads_at_risk*hazard_female_female)
```

We can see also that the wait time between female-female events is much lower. This is the case both because there are more females in the class and because males have a lower hazard of taking part in social interactions. 

Now, let's add our teacher variable to the model. This is accomplished by creating new CovSnd and CovRec matrices that include the teacher variable.

```{r}
CovSnd2=cbind(attributes[,c("intercept", "male", "teacher")])
CovRec2=cbind(attributes[,c("male", "teacher")])
```

And now we rerun our model with the updated CovSnd and CovRec matrices. 

```{r results='hide'}
mod2b=rem.dyad(edgelist_date1, n=class_size, effects=c("CovSnd", "CovRec"), 
                covar=list(CovSnd=CovSnd2, CovRec=CovRec2), ordinal=FALSE, hessian=TRUE)

```

Again, we can add better labels to the variable names and summarize the results. 

```{r}
coef.names=c("intercept", "Sender.male", "Sender.teacher", "Receiver.male", "Receiver.teacher")
names(mod2b$coef)=coef.names
summary(mod2b)
```

It looks like the teacher variables do not add much to the model. Let's compare the fit using BIC across the two models.

```{r}
mod2a$BIC - mod2b$BIC
```

Given that we want lower values, the simple model (mod2a, which just includes gender) would appear to be the better option. 

As a third kind of term, we will consider adding co-events to the model. Co-events are predictors that are based on attributes of a dyad. Here, we will add the seating structure of the class to the model. The basic idea is that nodes who are close in the classroom are more likely to talk to one another. Let's first read in the data: 

```{r}
seating_date1=read.table("https://sites.google.com/site/jeffreysmithdatafiles/class_seating_date1.txt", 
                         header=T)
head(seating_date1)
```

Here we set the size of the network on the edgelist: 

```{r}
attr(seating_date1,"n")=class_size
```

As in the previous tutorial, the data is stored as an edgelist, indicating if node i is sitting adjacent to node j. The rem.dyad function requires that this information be transformed into a matrix. So, we will go ahead and create a matrix of seating, where there is a 1 if i and j are next to each other in the class and 0 otherwise. We will accomplish this by taking the seating edgelist, turning it into a network object and then extracting the matrix from the network object. 

```{r}
seating_network_date1=network(x=seating_date1, directed=F) 
seating_matrix_date1=as.matrix(seating_network_date1)
seating_matrix_date1
```

Now, we can run our model. We must include "CovEvent" in the effects input. We must also add the seating matrix to the list of covariates (covar), set with CovEvent. We will use the CovSnd and CovRec matrices that only include gender (so no teacher variable).

```{r results='hide'}
mod3a=rem.dyad(edgelist_date1, n=class_size, effects=c("CovSnd", "CovRec", "CovEvent"), 
                covar=list(CovSnd=CovSnd1, CovRec=CovRec1, CovEvent=seating_matrix_date1), 
                ordinal=FALSE, hessian=TRUE)

```

Let's compare the fit between our previously preferred model and our new model. 

```{r}
mod2a$BIC - mod3a$BIC
```

It looks like the seating arrangement does strongly shape what events occur in the classroom, as the fit is dramatically improved. Now, let’s go ahead and add a second co-event matrix to the model. Here, we add information about the friendships that exist in the classroom. Friendship information was collected for each semester. Students were asked who they hung around with in the class. We will treat this information like a co-event, with the idea that interactions during the class period are more likely to involve friends than non-friends. We will first read in the data for friendship during the second semester (when the class of interest took place). 

```{r}
friends_sem2=read.table("https://sites.google.com/site/jeffreysmithdatafiles/class_edgelist_sem2.txt", 
                        header=T)
head(friends_sem2)
```

The edgelist captures if student i nominated student j as a friend. Note that the ids must match that found on the other data (interaction data, attributes, etc.). Here we set the size of the network on the edgelist: 

```{r}
attr(friends_sem2, "n")=class_size
```

As before we need to need to turn our edgelist into a matrix of 0s and 1s. 

```{r}
friends_sem2_network=network(x=friends_sem2, directed=T) 
friends_matrix_sem2=as.matrix(friends_sem2_network)
friends_matrix_sem2
```

While it is relatively simple to include a single co-event matrix in the model (see seating example above), it is a bit tricky to include multiple co-event matrices. The rem.dyad function requires that multiple matrices first be put together as an p X n X n array, where p is the number of matrices and n is the size of the network. 

We will go ahead and create that array. We will first create an array of NAs with the right structure (2 X 18 X 18).

```{r}
CovEvent_date1=array(data=NA, dim=c(2, class_size, class_size))
```

We will now put the first matrix, the seating matrix, in the first slot. 

```{r}
CovEvent_date1[1,,]=seating_matrix_date1
```

We will now put the second matrix, the friendship matrix, in the second slot. 

```{r}
CovEvent_date1[2,,]=friends_matrix_sem2
```

Checking the dimensions:

```{r}
dim(CovEvent_date1)
```

Note, that we would get unexpected (i.e., wrong) results if we had created the array to be 18 X 18 X 2. We are now in a position to run the model, putting in the newly created CovEvent array as input into the covar list. 

```{r results='hide'}
mod3b=rem.dyad(edgelist_date1, n=class_size, effects=c("CovSnd", "CovRec", "CovEvent"),
                covar=list(CovSnd=CovSnd1, CovRec=CovRec1, CovEvent=CovEvent_date1), 
                ordinal=FALSE, hessian=TRUE)
```

```{r}
summary(mod3b)
```

Here we look at the results after adding meaningful names to the coefficients:

```{r}
names(mod3b$coef)=c("Intercept", "Sender.male", "Receiver.male", "Adjacent Seating", "Friendship")
summary(mod3b)
```

We can see that both friendship and adjacent seating predict the occurrence of a social interaction event between i and j. Thus, if i and j sit next to each other and/or are friends they are much more likely to have an interaction event. Note that the gender differences in sending and receiving still seem to be present. 

#### 1.3. Micro Rules of Interaction
So far we have built up a simple but plausible model of interactions in this classroom. Girls tend to talk more than boys, while friends and those sitting close to each other also tend to interact during class. What we have yet to capture is something about the 'rules' of interaction. For example, we might expect turn taking (i talks to j and then j talks to i) above what we can capture from friendship and seating effects alone. 

We will build up our model slowly, including more complicated rules as we go along. As a start, let's include terms that capture recency of events. The two terms of interest are "RRecSnd" and "RSndSnd". With RRecSnd, we test if i is more likely to talk to j if j recently talked to i. With RSndSnd, we test if i is more likely to talk to j if i recently talked to j. The effects capture the idea that if j recently talked to i (or i recently talked to j), then an i->j event is more likely to be the next event. We will specify this model by including the "RRecSnd" and "RSndSnd" in the vector of effects. Note that we do not need to add anything to the covar list. For this first model we will not control for friendship or seating. 

```{r results='hide'}
mod4a=rem.dyad(edgelist_date1, n=class_size, effects=c("CovSnd", "CovRec", "RRecSnd", "RSndSnd"),
               covar=list(CovSnd=CovSnd1, CovRec=CovRec1), ordinal=FALSE, hessian=TRUE)

```

```{r}
summary(mod4a)
```

Adding meaningful names to the coefficients:

```{r}
names(mod4a$coef)=c("Recency.ji", "Recency.ij", "Intercept", "Sender.male", "Receiver.male")
summary(mod4a)
```

We see a positive coefficient on the recency receiver effects, suggesting that if j recently talked to i then i is likely to talk to j. On the other hand, there is a negative coefficient for the recency sender effects. This means that if i recently talked to j the next event is actually less likely to be i to j again. Note that the gender effects remain. Now we run the full model with seating and friendship included. 

```{r results='hide'}
mod4b=rem.dyad(edgelist_date1, n=class_size, effects=c("CovSnd", "CovRec", "CovEvent","RRecSnd", "RSndSnd"), 
               covar=list(CovSnd=CovSnd1, CovRec=CovRec1, CovEvent=CovEvent_date1), 
               ordinal=FALSE, hessian=TRUE)

```


```{r}
names(mod4b$coef)=c("Recency.ji", "Recency.ij", "Intercept", "Sender.male",
                    "Receiver.male", "Adjacent Seating", "Friendship")

summary(mod4b)
```

The first thing to note is that the model fit is much improved from our previously preferred model.

```{r}
mod3b$BIC - mod4b$BIC

```

The second thing to note is that the gender coefficients are no longer significant. This suggests that controlling for both recency of events and seating explains the gender differences in social interaction. Girls are more likely to sit next to each other; couple this with the interactional tendencies to respond to a recent interaction, and we see why girls are more likely to be involved in events.   

Now, let's see if we can consider other rules that may be shaping how nodes in this classroom interact with each other. We will now consider terms that capture p-shifts, or participation shifts. p-shifts are based strictly on the most recent event that occurred, rather than recency effects which can go back further in time. The idea is to capture micro rules in how interactions play out, based on the last interaction that took place. For our first example, we will add a turn taking rule, where A talks to B and the very next event is B talking to A. This is specified as PSAB-BA (as part of the effects vector). The rest of the model is the same as the previous model. 

```{r results='hide'}
mod4c=rem.dyad(edgelist_date1, n=class_size, 
               effects=c("CovSnd", "CovRec", "CovEvent", "RRecSnd", "RSndSnd", "PSAB-BA"), 
               covar=list(CovSnd=CovSnd1, CovRec=CovRec1, CovEvent=CovEvent_date1),
               ordinal=FALSE, hessian=TRUE)

```


```{r}
summary(mod4c)
```


```{r}
names(mod4c$coef)=c("Recency.ji", "Recency.ij", "Intercept", "Sender.male", 
                    "Receiver.male", "Adjacent Seating", "Friendship", "PSAB-BA")
summary(mod4c)
```

We can see that the fit is improved greatly, as there is a large effect of turn taking in social interactions. We can also see that the effects for recency are much reduced from the previous model. Now, let's add a somewhat more complicated interactional rule. Here we will add a term for 'turn continuing', PSAB-AY. This means that A talks with B and the very next event is A talking to someone else (besides B). We will also add a term for 'turn receiving', PSAB-BY. Here, A talks to B and the very next event is B talking to someone else (besides A).

```{r results='hide'}
mod4d=rem.dyad(edgelist_date1, n=class_size, 
               effects=c("CovSnd", "CovRec", "CovEvent", "RRecSnd", "RSndSnd", "PSAB-BA", "PSAB-BY", "PSAB-AY"), 
               covar=list(CovSnd=CovSnd1, CovRec=CovRec1, CovEvent=CovEvent_date1),
               ordinal=FALSE, hessian=TRUE)

```

```{r}
summary(mod4d)
```


```{r}
names(mod4d$coef)=c("Recency.ji", "Recency.ij", "Intercept", "Sender.male",
                    "Receiver.male", "Adjacent Seating", "Friendship", "PSAB-BA", "PSAB-BY", "PSAB-AY")

summary(mod4d)

```

Looking at our two added terms, we see that only the PSAB-BY coefficient is significant (at traditional levels). This suggest that there are norms about whose 'turn it is' to talk. Once A talks to B, it is now B's turn to talk. They are very likely to talk back to A (PSAB-BA) but may also talk to another node (PSAB-BY). There is little evidence for nodes talking twice in a row to different people.  It looks like the model fit is improved very slightly from the previous model:

```{r}
mod4c$BIC - mod4d$BIC
```

Let's look at one more set of terms, here focusing on interactional tendencies related to usurping the conversation. We include p-shift terms for PSAB-XA and PSAB-XB. With PSAB-XA, A talks to B and then another node (X) usurps the conversation and answers A. With PSAB-XB, A talks to B and then another node (X) usurps the conversation and talks to B. In both cases, a node talks 'out of turn' relative to the previous event. Let's go ahead and add these terms to the model. 

```{r results='hide'}
mod4e=rem.dyad(edgelist_date1, n=class_size,
                effects=c("CovSnd", "CovRec", "CovEvent", "RRecSnd", "RSndSnd",
                          "PSAB-BA", "PSAB-BY","PSAB-AY", "PSAB-XA", "PSAB-XB"), 
                covar=list(CovSnd=CovSnd1, CovRec=CovRec1, CovEvent=CovEvent_date1),
                ordinal=FALSE, hessian=TRUE)
```


```{r}
summary(mod4e)
```


```{r}
names(mod4e$coef)=c("Recency.ji", "Recency.ij", "Intercept", "Sender.male",
                    "Receiver.male", "Adjacent Seating", "Friendship",
                    "PSAB-BA", "PSAB-BY", "PSAB-XA", "PSAB-XB", "PSAB-AY")

summary(mod4e)

```

We can see that the PSAB-XB effect is much stronger than the PSAB-XA effect. This suggests that if one usurps the conversation from A to B, then one must interact with B in the next interaction, in essence giving B their rightful turn coming up next. And again, it looks like the model fit is improved a bit from the previous model. 

```{r}
mod4d$BIC - mod4e$BIC
```

So, the overall story is one where there are clear micro rules to interacting in a classroom (above the effects for friendship, seating and so on). The basic rules could be summarized as: if A talks to B then the next event should be B to A; B to someone else; or someone else to B. Thus, there are clear norms about turn taking. If A talks to B, B is very likely to be part of the next event (one way or another). Of course, we could imagine looking at other kinds of terms, but this is pretty good start to the model. 

#### 1.4. Assessing Model Fit
In interpreting our results, it is useful to see if our models are fitting well. While BIC can offer some evidence if one model is preferred to another, we can look at the residuals and the predicted classification to see how well the model is actually predicting the data. Here we will work with mod4e, our preferred model from above. One useful part of the output is predicted.match. 

```{r}
head(mod4e$predicted.match)
```

Each row corresponds to an observed event. The first column shows if the model predicted the sender of that event correctly and the second column shows if the model predicted the receiver of that event correctly. Note that the model is trying to predict the specific sequence of events (i.e., the exact order of sender-receiver events). Let's see how we did by doing a table of the send and receive columns. 

```{r}
table(mod4e$predicted.match[,"send_col"], mod4e$predicted.match[,"receive_col"])
```

We can see that 119 times we predicted the exact sender and the exact receiver correct (in sequence), while 89 times we got neither the sender nor the receiver correct. And now let's transform the table into proportions, showing the proportion where we get the exact sequence right (wrong, etc.).

```{r}
prop.table(table(mod4e$predicted.match[,"send_col"], mod4e$predicted.match[,"receive_col"]))
```

We can see that about 52% of the time we get the exact event correct, while 39% of the time we miss completely and do not get the sender or receiver. The model is thus doing an okay job of prediction but is clearly missing some element that is important for predicting interaction events. Let's see if we can identify the cases (i.e., events) where the model is not doing such a good job at prediction. We first summarize the residuals for the model. 

```{r}
summary(mod4e$residuals)
```

Now, we will identify some of the outlying cases, those with high residuals. We will define that for convenience as cases with residuals greater than 10. 

```{r}
high_resids=which(mod4e$residuals>10)
```

Here we take a look at the events with high residuals, reducing the edgelist to just those cases where the model is not fitting well. 

```{r}
edgelist_date1[high_resids,]
```

Node 12 seems to show up quite a bit in the receiver column of these events. Let's take a look at the attributes for node 12. 

```{r}
attributes[12, ]
```

Node 12 is the teacher in the class. So, perhaps we were too hasty in removing the teacher variable, as it looks like we are missing the set of interactions where students talk socially to the teacher. Let’s rerun our model but use CovSnd2 and CovRec2, which includes both the gender and teacher effects. 

```{r results='hide'}
mod4f=rem.dyad(edgelist_date1, n=class_size,
                effects=c("CovSnd", "CovRec", "CovEvent", "RRecSnd", "RSndSnd",
                          "PSAB-BA", "PSAB-BY","PSAB-AY", "PSAB-XA", "PSAB-XB"), 
                covar=list(CovSnd=CovSnd2, CovRec=CovRec2, CovEvent=CovEvent_date1),
                ordinal=FALSE, hessian=TRUE)

```

Let's check the fit compared to the previous model (with no teacher sender/receiver effects).

```{r}
summary(mod4f)
mod4e$BIC - mod4f$BIC
```

It looks like adding the teacher terms did help the fit a bit. And now let's look at the results. 

```{r }
names(mod4f$coef)=c("Recency.ji", "Recency.ij", "Intercept", "Sender.male", "Sender.teacher", 
                    "Receiver.male", "Receiver.teacher", "Adjacent Seating", "Friendship", 
                    "PSAB-BA", "PSAB-BY", "PSAB-XA", "PSAB-XB", "PSAB-AY")

summary(mod4f)
```

We can see that the teacher is part of more interactions as the receiver (being talked to) than we would expect based on other terms in the model. This is likely the case because the teacher can easily talk to anyone in the class (i.e., the teacher is not subject to only talking to those adjacent to them in the classroom), and so the term on the seating arrangement pushes the previous model to under predict interactions with the teacher. This is now rectified in the current model. 

#### 1.5. Comparison to a Second Date (Optional Second Example)
We have so far run a number of models, interpreted the results and learned a bit about the interactional dynamics in this classroom. Here, we run through the same exercise (an abbreviated version) using interactional data from a different date. The classroom is the same, so the actors are the same, but this class takes place later in the second semester. More importantly, this was a date where there was a great deal more misbehaving in the class and the teacher had to sanction students to a much larger extent. Our question is how (or if) the interactional tendencies are different in a day where the class is less orderly and controlled. We begin by reading in the interactional data for this second date. 

```{r}
interactions_date2=read.table("https://sites.google.com/site/jeffreysmithdatafiles/class_interactions_date2.txt", 
                              header=T)
```

And again, we will take out those interactions where one node talks to the entire class simultaneously. 

```{r}
interactions_date2=interactions_date2[interactions_date2$to_all_col==0 & interactions_date2$from_all_col==0,]
```

And now we create the edgelist matrix, adding a row for the stop time for the interactions (again, .10 standardized minutes after the last recorded interaction). 


```{r}
edgelist_date2=as.matrix(interactions_date2[,c("time_estimate_col", "send_col", "receive_col")])
tail(edgelist_date2)
edgelist_date2=rbind(edgelist_date2, c(33.10, NA, NA))
```

The friendship data is the same as above (corresponding to the second semester) but we need to read in the seating data for this day. 

```{r}
seating_date2=read.table("https://sites.google.com/site/jeffreysmithdatafiles/class_seating_date2.txt", 
                         header=T)
head(seating_date2)
```


Here we turn the edgelist into a matrix, as we did before. 

```{r}
seating_network_date2=network(x=seating_date2, directed=F) 
seating_matrix_date2=as.matrix(seating_network_date2)
```

And once again, we need to create a covEvent array with the new seating matrix. 

```{r}
CovEvent_date2=array(data=NA, dim=c(2, class_size, class_size))
```

We will now put the first matrix, the seating matrix, in the first slot. 

```{r}
CovEvent_date2[1,,]=seating_matrix_date2

```

We will now put the second matrix, the friendship matrix, in the second slot. 

```{r}
CovEvent_date2[2,,]=friends_matrix_sem2
```

We are now in a position to run the same model as we did above. We will just run the preferred model (mod4f), with all terms included. Note that the CovRec and CovSnd matrices are the same as above (as the attributes are the same across time in this case).

```{r results='hide'}
mod4f_date2=rem.dyad(edgelist_date2, n=class_size,
                effects=c("CovSnd", "CovRec", "CovEvent", "RRecSnd", "RSndSnd",
                          "PSAB-BA", "PSAB-BY","PSAB-AY", "PSAB-XA", "PSAB-XB"), 
                covar=list(CovSnd=CovSnd2, CovRec=CovRec2, CovEvent=CovEvent_date2),
                ordinal=FALSE, hessian=TRUE)
```

Let's add some more meaningful variable names. In, this case the terms are the same as with mod4f, so we can grab those names and use them directly. 

```{r}
names(mod4f_date2$coef)=names(mod4f$coef)
summary(mod4f_date2)
```

We will create a little data frame to compare the coefficients from our two days. 

```{r}
compare_coefs=data.frame(date1=mod4f$coef, date2=mod4f_date2$coef)
compare_coefs
```

Overall, much of the same interactional rules we saw above (in the 'normal' day) hold when looking at this second date, where the class was more unruly. We still see turn taking in interactions, for example (AB and then BA). We still see rules around usurping the conversation, such that when A talks to B and then X jumps in, they are likely to talk to B. Still, there would appear to be some potentially important differences (we would want to explore this more formally). For example, the effects for friendship and seating are particularly important for the second day. Similarly, there is some evidence that the tendency for PSAB-BY p-shifts are relatively high here. A class that is more unruly tends to have interactions that are based more on friendship and adjacent seating (i.e. talking to neighbors rather than doing discussion). Additionally, there may be a higher (relative) tendency for nodes to form a kind of two-step interaction (A-B-Y) rather than just a simple return to the person addressing them (A-B-A). This would potentially create more disruption in the classroom, as a larger number of students are brought into the initial interaction event. 

It is also useful to compare these results to the kind of models we saw in the previous tutorial on STERGM. In general, STERG models allow us to test hypotheses about formation and persistence of ties. This opens up questions about triadic effects and larger group structures. In the language of relational event models, A may talk to B, and then B may talk to Y; but  when B talks to Y, Y is very likely to be someone that A generally talks with. Thus, little groups in the classroom emerge that are harder to see in the relational event model than with STERGM. On the other hand, STERGM completely obscure the actual dynamics of moment-to-moment interactions, missing some of the ‘rules’ of interaction that come out so clearly in the relational event results. 

Chapter 13 has covered statistical network models, moving from the cross-sectional case all the way up to continuous-time network data. In Chapter 14, we still  utilize statistical network models, but we focus on problems related to peer influence and diffusion. 


